**Test a Perceptual Phenomenon by Adam McDonald**
========================================================

The Stroop Effect comes from an experiment that tries to show the effect
of interference on the reaction time of a task. In the test people are shown a 
set of names of colors printed in their respective color (e.g., the word red 
printed in red). The time it takes to read the set aloud is recorded. Then they
are shown a set of names of colors printed in colors different than the name
(e.g., the word red printed in blue). The time to read this set is recorded. 
In theory the congruent set will take less time to read than the incongruent
set because the different colors in the incongruent set will cause 
interference and slow downs reaction times.

## Variables
In this experiment, the independent variable is which set is read aloud, 
congruent or incongruent. The dependent is the length of time that it takes the 
person to read the set of words aloud. 

##Hypotheses
The test will be a one-sided paired t test. It will be one-sided because I am 
trying to determine if the time to read the incongruent set will be 
significantly slower than the time to read the congruent set 
Since the two sets were given to the same people to read, it will be a paired
test. Since we only have samples and don't know population statistics the test
will be a t-test. As we will see below in the plots, the distributions of
the samples are normal and we assume the population distribution would be normal
as well. The size of the sample is large enough at 24, and the variance of each
sample is relatively close. Based on these assumptions the use of a t-test is
valid.

The null hypothesis is that the incongruent population average time to read is
less than or equal to the congruent population average time to read.  
H~0~: $\mu$~Incongruent~ <= $\mu$~Congruent~

The alternative hypothesis is that the incongruent population average time to 
read is greater than the congurent population average time to read.  
H~A~: $\mu$~Incongruent~ > $\mu$~Congruent~

## Dataset

The dataset consists of 24 observations with the length of time it took to
read both the congurent and incongruent sets I have also included the 
difference between the 2 times. 

```{r echo=FALSE, message=FALSE, warning=FALSE, setup}
library(ggplot2)
library(gridExtra)
df <- read.csv('StroopData.csv')
print(df)
```

## Descriptive Statistics

**Congruent Sample**  
$\bar{x}$~Congruent~ = 14.05  
sd~Congruent~ = 3.56

**Incongruent Sample**  
$\bar{x}$~Incongruent~ = 22.02  
sd~Incongruent~ = 4.80

## Histograms of Congruent and Incongruent samples

```{r echo=FALSE,message=FALSE,warning=FALSE,fig.width=10,fig.height=10,plots}
p1 <- ggplot(aes(x=Congruent),data=df) +
  xlab('Congruent Sample (seconds)') +
  geom_histogram(binwidth=1,fill = "red",color='black',alpha=.4) +  
  geom_vline(aes(xintercept = mean(Congruent)),size = 1, colour = "red",
                   linetype = "dashed") + xlim(8,36)

p2 <- ggplot(aes(x=Incongruent),data=df) +
  xlab('Incongruent Sample (seconds)') +
  geom_histogram(binwidth=1,fill = "blue",color='black',alpha = 0.4) +
  geom_vline(aes(xintercept = mean(Incongruent)),size = 1, colour = "blue",
                   linetype = "dashed") + xlim(8,36)

grid.arrange(p1, p2)
```

As seen in the histograms both of the samples are relatively normal, but the
inconguent sample is shifted to the right of the x-axis. The t test will
show if the shift right is statistically significant.

## One-Sided Paired t test

$\alpha$ = .05  
t critical = -1.71  
n = 24  
df = 23  
$\bar{x}$~Congruent-Incongruent~ = -7.96  
s = 4.86  
se = .99  
t = -8.02  
p < .0005

Using confidence level of 95% and an $\alpha$ of .05 the t critical value is 
-1.71. The t statistic from the test of -8.02 is much lower than the t critical
value, which means that the null hypothesis is rejected. The difference between 
the means of -7.96 is statistically significant. This can also be seen by 
comparing the $\alpha$ of .05 to the p-value. Using the t table the p-value is
less than .0005, which means that it is less than the $\alpha$.

As expected changing the color used to print the color name does have an
effect on reaction time. When the printed color is different than the color 
name it takes more time to read the word aloud.