**Prosper Loan Data Exploration by Adam McDonald**
========================================================

# Overview of Dataset

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(GGally)
library(scales)
library(memisc)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}

loanData <- read.csv('prosperLoanData.csv')

```

The dataset includes 113,937 loans with 81 variables. 

```{r echo=FALSE, message=FALSE, warning=FALSE, dim}

dim(loanData)

```
  
The variables are:

```{r echo=FALSE, message=FALSE, warning=FALSE, variables}

names(loanData)

```
  
I will limit the data to only the below 10 variables that I would like to 
explore further.

```{r echo=FALSE, message=FALSE, warning=FALSE, filter_dataframe}

df <- loanData[c("LoanStatus", "BorrowerAPR", "EstimatedReturn", 
                 "ProsperRating..numeric.", "EmploymentStatus", 
                 "CreditScoreRangeLower", "DebtToIncomeRatio", "IncomeRange", 
                 "LoanOriginationDate", "MonthlyLoanPayment")]

names(df)
summary(df)

```

# Univariate Plots Section

First I will look at variables related to the loans:

### Loan Origination Date

```{r echo=FALSE, message=FALSE, warning=FALSE, loanoriginationdate}

summary(as.Date(df$LoanOriginationDate))

ggplot(aes(x = substr(LoanOriginationDate, 1, 4)), data = df) +
  geom_histogram(fill = "steelblue", color = "black", stat = "count") +
  xlab("Loan Origination Year") +
  ylab("Count")

```

The data includes loans from November 2005 through March 2014. 
There was a low point during the financial crisis in 2009, but we see
solid growth in the number of loans each year from there.

### Loan Status

```{r echo=FALSE, message=FALSE, warning=FALSE, loanstatus}

summary(df$LoanStatus)

ggplot(aes(x = LoanStatus), data = df) +
  geom_bar(aes(y = (..count..)/sum(..count..)*100),
           fill = "steelblue", color = "black", stat = "count") +
  ylab("Percent of Total") +
  xlab("Loan Status") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

About 15% of loans fall into either Charged Off or Defaulted, while just under 
85% or Completed or Current. The remaining items are small amounts of the total
number of loans. I will convert the Loan Status into a column that is either
Performing or Non-Performing. The performing will be either completed or 
current and the non-perfomring will be anything else.

### Loan Status

```{r echo=FALSE, message=FALSE, warning=FALSE, loanstatus2}

df$LoanStatus <- ifelse(df$LoanStatus == 'Completed' | df$LoanStatus == 
                          'Current', 'Performing', 'Non-Performing')

summary(df$LoanStatus)

ggplot(aes(x = LoanStatus), data = df) +
  geom_bar(aes(y = (..count..)/sum(..count..)*100),
           fill = "steelblue", color = "black", stat = "count") +
  ylab("Percent of Total") +
  xlab("Loan Status") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

### Monthly Loan Payment

```{r echo=FALSE, message=FALSE, warning=FALSE, MonthlyLoanPayment}

summary(df$MonthlyLoanPayment)

ggplot(aes(x = MonthlyLoanPayment), data = df) +
  geom_histogram(fill = "steelblue", color = "black", binwidth = 50) +
  ylab("Count") +
  xlab("Monthly Loan Payment")+
  geom_vline(aes(xintercept = mean(MonthlyLoanPayment)),col='blue',size=2)

```

This is right skewed with the average amount aroud $270 per month and 
the median around $217 per month. Most loans
have a pretty small monthly payment amount. 
 
 
 
Now I will shift to variables that are related to the borrower:

### Employment Status

```{r echo=FALSE, message=FALSE, warning=FALSE, EmploymentStatus}

summary(df$EmploymentStatus)

ggplot(aes(x = EmploymentStatus), data = df) +
  geom_histogram(fill = "steelblue", color = "black", stat = "count") +
  xlab("Employment Status") +
  ylab("Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

Borrowers are mostly employed, but there is a small amount not employed or 
retired.

### Income Range

```{r echo=FALSE, message=FALSE, warning=FALSE, IncomeRange}

summary(df$IncomeRange)

ggplot(aes(x = IncomeRange), data = df) +
  geom_histogram(fill = "steelblue", color = "black", stat = "count") +
  xlab("Income Range") +
  ylab("Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

Looks like the factors are out of order for the Income Range, I will adjust 
this and redisplay the graph.

```{r echo=FALSE, message=FALSE, warning=FALSE, IncomeRange2}

df$IncomeRange <- factor(df$IncomeRange, levels = c("$0","$1-24,999",
                                                    "$25,000-49,999",
                                                    "$50,000-74,999",
                                                    "$75,000-99,999",
                                                    "$100,000+","Not displayed",
                                                    "Not employed"))
ggplot(aes(x = IncomeRange), data = df) +
  geom_histogram(fill = "steelblue", color = "black", stat = "count") +
  xlab("Income Range") +
  ylab("Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

The income range isn't quite normally distributed. 
It appears a little off, with $100,000 being somewhat high. This could be 
because the income range is self reported or it could just be higher income 
people are using more loans.

### Debt to Income Ratio

```{r echo=FALSE, message=FALSE, warning=FALSE, DebtToIncomeRatio}

summary(df$DebtToIncomeRatio)

ggplot(aes(x = DebtToIncomeRatio), data = df) +
  geom_histogram(fill = "steelblue", color = "black",stat = "count") +
  xlab("Debt to Income Ratio") +
  ylab("Count")

```

This is skewed by outliers, and tough to see the distribution.
I am going to exclude the top 1% from the graph.

```{r echo=FALSE, message=FALSE, warning=FALSE, DebtToIncomeRatio2}

ggplot(aes(x = DebtToIncomeRatio), data = subset(df,!is.na(DebtToIncomeRatio)))+
  geom_histogram(fill = "steelblue", color = "black",stat = "count") +
  xlab("Debt to Income Ratio") +
  ylab("Count") +
  xlim(limits=c(0,quantile(
    subset(df,!is.na(DebtToIncomeRatio))$DebtToIncomeRatio,.99))) +
  geom_vline(aes(xintercept = mean(DebtToIncomeRatio)),col='blue',size=2)
```

The distribution is slightly right skewed, The mean is 0.276, while the median is
only 0.22.

### Credit Score

```{r echo=FALSE, message=FALSE, warning=FALSE, CreditScoreRangeLower}

summary(df$CreditScoreRangeLower)

ggplot(aes(x = CreditScoreRangeLower), data = 
         subset(df, df$CreditScoreRangeLower > 0.0)) +
  geom_histogram(fill = "steelblue", color = "black", binwidth = 10) +
  xlab("Credit Score") +
  ylab("Count") +
  geom_vline(aes(xintercept = mean(CreditScoreRangeLower)),col='blue',size=2)

```

The distribution of credit scores is centered at a median of 680 and a mean
or 686. It is barely skewed left with lower credit scores most likely not 
being approved as frequently.

### Prosper Rating

```{r echo=FALSE, message=FALSE, warning=FALSE, ProsperRating}

summary(df$ProsperRating..numeric.)

ggplot(aes(x = factor(ProsperRating..numeric.)), data = df) +
  geom_bar(fill = "steelblue", color = "black") +
  xlab("Prosper Rating") +
  ylab("Count")

```

The Prosper Rating distribution is relatively normal.
This makes sense as they are rating the risk of loans. They don't necessarily
want too many highly rated loans because those don't provide much of a return,
but the higher return on riskier loans doesn't necessarily outwiegh the risk
of default.

### Borrower APR

```{r echo=FALSE, message=FALSE, warning=FALSE, BorrowerAPR}

summary(df$BorrowerAPR)

ggplot(aes(x = BorrowerAPR), data = subset(df,!is.na(BorrowerAPR))) +
  geom_histogram(fill = "steelblue", color = "black", binwidth =  .01) +
  xlab("Borrower APR") +
  ylab("Count") +  
  geom_vline(aes(xintercept = mean(BorrowerAPR)),col='blue',size=2)

```

The APR distribution is unimodal, but has an outlier peak around .375. 
The mean is .20976 and median is .21883. 
They could just use a higher rate for many risky loans to offset the risk of
default.


### Estimated Return

```{r echo=FALSE, message=FALSE, warning=FALSE, EstimatedReturn}

summary(df$EstimatedReturn)

ggplot(aes(x = EstimatedReturn), data = subset(df,!is.na(EstimatedReturn))) +
  geom_histogram(fill = "steelblue", color = "black", binwidth = .005) +
  xlab("Estimated Return") +
  ylab("Count") +  
  geom_vline(aes(xintercept = mean(EstimatedReturn)),col='blue',size=2)

```

The estimated return follows a similar distribution as the APR, which makes 
sense because the interest rate is set to give a return based on the amount 
of risk involved in the loan.

# Univariate Analysis

### What is/are the main feature(s) of interest in your dataset?
I am most interested in the interest rate for the loans. I am interested to see
what impacts the interest rates the most and how it relates to the estimated
return.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
I am interested in seeing if the Loan Status can be predicted by any certain
variable. For example, do lower Prosper Ratings lead to higher non-performance 
rates? I am also interested to see which variables influence interest rates
the most.

### Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
I changed the order of factors in Income Range so that they showed in order
when creating the histogram. I also changed Loan Origination Date to a date 
format instead of a factor. Finally, I changed the Loan Status to be either
performing or non-performing to make looking at the status a little easier.

# Bivariate Plots Section

```{r echo=FALSE,message=FALSE,warning=FALSE,fig.width=15,fig.height=20,ggpair}
df$LoanOriginationDate <- as.Date(df$LoanOriginationDate , "%Y-%m-%d")

columnnames = c('Loan Status','Borrower APR','Estimated Return','Prosper Rating',
                'Employment Status','Credit Score','Debt to Income',
                'Income Range','Loan Orig Date','Monthly Loan Payment')

ggpairs(df,columnLabels = columnnames,
        labeller = label_wrap_gen(10), axisLabels = 'none')
```

The highest correlation appears to be between the Prosper Rating and the 
Borrower APR. Next is the Estimated Return and Interest Rate followed by the 
Estimated Return and Prosper Rating. This makes sense, the Interest Rate is 
based off the Prosper Rating and the Estimated Return is mostly based off the 
Prosper Rating, but also from the size and length of the loan.

I will begin by looking at the Loan Status and different variables.

### Prosper Rating by Loan Status

```{r echo=FALSE, message=FALSE, warning=FALSE, prosperratingboxplot}
ggplot(aes(y = ProsperRating..numeric., x = LoanStatus), data = df) +
  geom_boxplot()
```

Here we see that the higher the Prosper Rating the more likely the loan is 
to be performing. It seems that the Prosper Rating is pretty good at predicting
how risky the loan is. The median for Non-Performing loans is 3 and the median
of performing loans is 4.

### Credit Score by Loan Status

```{r echo=FALSE, message=FALSE, warning=FALSE, creditscoreloanstatus}
ggplot(aes(y = CreditScoreRangeLower, x = LoanStatus), data = 
         subset(df, df$CreditScoreRangeLower > 0.0)) +
  geom_boxplot()
```

It looks like the Credit Score is not quite as good at predicting if a loan
will be completed or not. As seen above the correlation between Credit Score
and Prosper Rating is only -0.346. There are other factors that go into the 
Prosper Rating that help predict Loan Status. The median score of 
Non-Performing loans is about 660 while Performing loans have a median
score just above 700.

### Income Range by Loan Status

```{r echo=FALSE, message=FALSE, warning=FALSE, incomerangeloanstatus}
ggplot(aes(x = IncomeRange, fill = LoanStatus), data = 
         subset(df, df$CreditScoreRangeLower > 0.0)) +
  geom_histogram(stat='count') +
  scale_fill_brewer(palette="Blues") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Income Range definitely doesn't predict if a loan will be completed or not. 
The percentage of non-performing loans doesn't increase with an increase in 
income.

### Loan Origination Date by Loan Status

```{r echo=FALSE, message=FALSE, warning=FALSE, origdateloanstatus}
ggplot(aes(y = LoanOriginationDate, x = LoanStatus), data = df) +
  geom_boxplot()
```

Over time a greater amount of loans have completed or are current. It would be
interesting to see if that is because people taking loans have better credit 
scores some other factor. The median of Non-Performing loans is late 2008,
while the median of Performing loans is late 2012.

Now I will take a look at the relationship of Borrower APR with different
variables.

### Prosper Rating by Borrower APR

```{r echo=FALSE, message=FALSE, warning=FALSE, APRprosperrating}
ggplot(aes(y = BorrowerAPR, x = factor(ProsperRating..numeric.)), data = 
         subset(df, !is.na(ProsperRating..numeric.))) +
  geom_boxplot()
```

As the Prosper Rating increases the Borrower APR charged goes down, which makes
sense because the less risky the loan the less interest borrowers have to pay.
The boxplots show  that the median APR goes down with each increase to the 
prosper rating.
As we saw above the correlation is -0.962 between these 2 variables.
It also looks like the variances of rates goes down as the Prosper Rating 
increases. This may be because any Prosper Rating lower than 4 takes more 
factors into account when choosing an interest rate.

### Estimated Return by Borrower APR

```{r echo=FALSE, message=FALSE, warning=FALSE, estimatedreturnAPR}
ggplot(aes(y = EstimatedReturn, x = BorrowerAPR), data = df) +
  geom_point() + geom_smooth(method='lm')
```

The Estimated Return and Borrower APR are positively correlated at 0.794. There
is definitely a relationship that as the rate increases the return does as well.
There is another factor here that is causing the estimated return to shift
up or down, we will explore this further in the Multivarate Analysis.

### Prosper Rating by Estimated Return

```{r echo=FALSE, message=FALSE, warning=FALSE, estimatedreturnprosperrating}
ggplot(aes(y = EstimatedReturn, x = factor(ProsperRating..numeric.)), data = 
         subset(df, !is.na(ProsperRating..numeric.))) +
  geom_boxplot()
```

It looks likes Prosper attempts to keep their estimated return about the same
across the different ratings. There is a -0.66 correlation between the 2 
variables that seems to be driven by to 0 rated loans having the much lower 
estimated returns.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
First I looked at the relationship between Loan Status and a few other 
variables. It appears that the Prosper Rating does show an ability to predict 
if a loan will complete or not. Some of the factors that go into the Prosper 
Rating like the Credit Score and Income Range by themselves aren't a good 
indicator as to whether a loan will complete or not. 
I also looked at the relationships between Borrower APR, Estimated Return, and
Prosper Rating. These are all pretty strongly correlated and have the closest
relationships of any variables in the dataset. 

### What was the strongest relationship you found?
The highest correlation was between Prosper Rating and Borrower APR. It was a
0.962 correlation, which makes sense, because they probalby use the rating to
determine what the APR will be for each loan.

# Multivariate Plots Section

### Estimated Return by Borrower APR and Prosper Rating

```{r echo=FALSE, message=FALSE, warning=FALSE, estimatedreturnprosperratingAPR}
ggplot(aes(y = EstimatedReturn, x = BorrowerAPR, color = 
             factor(ProsperRating..numeric.)), data = df) +
  scale_color_brewer(palette="Blues") + theme_dark() +
  geom_point() + geom_smooth(method='lm',color='red')
```

Here I look again at Estimated Return by Borrower APR, but add in a color
for the Propser Rating. This helps explains the banding that occurs. Yes
the Estimated Return does go up as the APR increases, but the line
is shifted up or down based on the Prosper Rating. The higher the Prosper
Rating the higher the floor of the Estimated Return is.

### Estimated Return by Borrower APR and Loan Origination Date

```{r echo=FALSE, message=FALSE, warning=FALSE,estimatedreturnprosperratingdate}
ggplot(aes(y = EstimatedReturn, x = BorrowerAPR, color = 
             LoanOriginationDate), data = df) +
  geom_point()  + geom_smooth(method='lm',color='red')
```

This is pretty interesting. It looks like the slope of the Estimated Return
vs Borrower APR has gotten lower over time. For more recent loans the 
increase in the interest rate doesn't increase the estimated return as much 
as it used to.

### Prosper Rating by Loan Status and Borrower APR

```{r echo=FALSE, message=FALSE, warning=FALSE, prosperratingloanstatusAPR}
ggplot(aes(y = ProsperRating..numeric., x = LoanStatus, color = 
             BorrowerAPR), data = df) +
  geom_jitter()
```

This plot shows that the higher the Prosper Rating the lower the percentage
of Non-Performing loans and the lower the APR. What is somewhat interesting
is there are some low Prosper Rating that have a higher APR, but are still
current or completed. Looking it appears there are more of those loans in
the Performing status.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
The Estimated Return, Prosper Rating, and Borrower APR are all three closely
related. There is a strong positive linear relationship between the interest
rate and the return. This linear relationship is shifted up or down based on 
the Prosper Rating.

### Were there any interesting or surprising interactions between features?
I was surprised that the slope of the Estimated Return vs Borrower APR has 
gotten lower over time. It seems that they have made some changes to make the
estimated return less variable regardless of the interest rates.


------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(aes(y = ProsperRating..numeric., x = LoanStatus), data = df) +
  geom_boxplot() +
  ylab("Prosper Rating (numeric 0-7)") +
  xlab("Loan Status") +
  ggtitle("Prosper Rating by Loan Status")
```

### Description One

I like this simple boxplot because it clearly shows that the Prosper Rating
historically has done a good job of rating the riskiness of a loan. Non-Performing
loans have had lower Prosper Ratings assigned at the origination of the loan.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(y = EstimatedReturn, x = BorrowerAPR, color = 
             factor(ProsperRating..numeric.)), data = df) +
  geom_point() + scale_color_brewer(palette="Blues") + theme_dark() +
  ylab("Estimated Return (percentage)") +
  xlab("Borrower APR (percentage)") +
  ggtitle("Estimated Return by Borrower APR and Prosper Rating") +
  labs(color = "Prosper Rating")  + geom_smooth(method='lm',color='red')
```

### Description Two

This plot does a good job of illustrating the relationship between Estimated
Return, Borrower APR, and Prosper Rating. There is a strong positive linear 
relationship between the interest rate and the return. This linear relationship
is shifted up or down based on the Prosper Rating. Including both
APR and Prosper Rating in a model could probably predict Estimated Return
pretty well.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(y = EstimatedReturn, x = BorrowerAPR, color = 
             LoanOriginationDate), data = df) +
  geom_point() +
  ylab("Estimated Return (percentage)") +
  xlab("Borrower APR (percentage)") +
  ggtitle("Estimated Return by Borrower APR and Loan Origination Year") +
  labs(color = "Loan Origination Year")  + geom_smooth(method='lm',color='red')
```

### Description Three

This plot is similar to the second plot, but changes out the Prosper Rating for
the Loan Origination Year. It again shows the relationship between Estimated
Return and Borrower APR, but shows how it is changing over time. In later years
the slope of the line decreases, meaning an increase in APR in 2014 increases
the Estimated Return less than it it did in 2006.

------

# Reflection

Using the Prosper loan data set, I selected 10 of the 81 variables to explore
further. I wanted to investigate the Loan Status and adjusted the variable 
to be Performing or Non-Performing to find how many loans completed or are 
current versus the other categories. I wanted to look and see which factors 
led to a loan not completing. I also looked at the Borrower APR, Estimated
Return, and Prosper Score. I wanted to see the relationship between the 3.

One of the difficuties was trying to choose which variables to look at. With 81
different variables it was necessary to narrow down my exploration. It would 
be interesting to go back and look at other variables in the data set knowing
what I do about the 10 I chose.

For future Analysis I think it would be interesting to create a model that 
usese variables to predict the Prosper Rating. I would then change the model
and see if I could create a rating that is even better predicting how a loan
will perform.